---
title: "Gaussian process demo with genomic data"
author: "Estee Y Cramer"
date: "4/29/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load Libraries
```{r}
library(cmdstanr) 
library(posterior)
options(pillar.neg = FALSE, pillar.subtle=FALSE, pillar.sigfig=2)
library(tidyr) 
library(dplyr) 
library(ggplot2)
library(bayesplot)
library(lubridate)

theme_set(bayesplot::theme_default(base_family = "sans", base_size=16))
set1 <- RColorBrewer::brewer.pal(7, "Set1")
SEED <- 48927 # set random seed for reproducability
```


# Load Data
```{r}
national_level_raw <- read.csv("national_level.csv") #file of national level human genomic data samples. Filtered in the file "filtering genomic file.RMD"
``` 


# Clean Data

## View data summaries

most of the variants are delta, omicron, or alpha
```{r}
national_level_raw %>% 
  group_by(Nextstrain_clade) %>%
  summarise(n = n()) %>% 
  arrange(-n)
```


There are 116 weeks worth of data 
```{r}
#filter missing dates
 national_level_raw %>% 
  mutate(year = year(date)) %>%
  mutate(epiweek = epiweek(date)) %>%
  group_by(year, epiweek) %>%
  summarise(n = n()) 
```

## Edit data file 

remove missing dates 
```{r}
clean_df <- national_level_raw %>%
  mutate(year = year(date)) %>%
  mutate(epiweek = epiweek(date)) %>%
  filter(!is.na(epiweek)) %>% #filter out rows with no date 
  mutate(epiweek_year = paste(year, formatC(epiweek, width=2, flag="0"), sep="_"))
```

Add binary column for 21K (Omicron) yes/no

If sample is omicron, binary_omicron = 1
```{r}
clean_df <- clean_df %>% 
  mutate(binary_omicron = ifelse(Nextstrain_clade == "21K (Omicron)", 1, 0))
```


## Simplify data
```{r}
omicron_date <- clean_df %>%
  select(binary_omicron, epiweek_year)
```

## Calculate % of data that is Omicron each week
```{r}
omicron_date_perc <- omicron_date %>%
  group_by(epiweek_year) %>%
  summarise(perc_omicron = round(sum(binary_omicron) / n(),4)) %>%
  mutate(num_week = row_number())
```



# Visualize data
```{r}
omicron_date_perc %>%
  ggplot(aes(x=epiweek_year, y=perc_omicron))+
  geom_point()+
  labs(x="Epi Week", y="Percent Omicron") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


# Stan Models

GP model with Hilbert basis functions
```{r}
file1 <- "gpbf1.stan"
writeLines(readLines(file1))
```

code includes Hilbert space basis function helpers
```{r}
writeLines(readLines("gpbasisfun_functions.stan"))
set_cmdstan_path()
```

# Build STAN model
```{r}
# set_cmdstan_path()
# install_cmdstan()
model1 <- cmdstan_model(stan_file = file1, include_paths = ".")
```


Pass data to STAN 
```{r}
standata1 <- list(x=omicron_date_perc$num_week,
                  y=omicron_date_perc$perc_omicron,
                  N=length(omicron_date_perc$num_week),
                  #didn't change any of the below from the example.
                  c_f=1.5, # factor c of basis functions for GP for f1. 
                  M_f=40,  # number of basis functions for GP for f1
                  c_g=1.5, # factor c of basis functions for GP for g3
                  M_g=40)  # number of basis functions for GP for g3

```


Sample using dynamic HMC
```{r}
fit1 <- model1$sample(data=standata1, iter_warmup=500, iter_sampling=500,
                      chains=4, parallel_chains=2, adapt_delta=0.9)
```

```{r}
draws1 <- fit1$draws()
summarise_draws(subset(draws1, variable=c('intercept','sigma_','lengthscale_'), regex=TRUE))
```

compare model to data
```{r}
draws1m <- as_draws_matrix(draws1)
Ef <- colMeans(subset(draws1m, variable='f'))
sigma <- colMeans(subset(draws1m, variable='sigma'))
pred<-data.frame(Ef=Ef,sigma=sigma)
cbind(omicron_date_perc, pred) %>%
  ggplot(aes(x=num_week,y=perc_omicron))+
  geom_point()+
  labs(x="Week (number)", y="Percent Omicron")+
  geom_line(aes(y=Ef), color=set1[1])+
  geom_line(aes(y=Ef-2*sigma), color=set1[1],linetype="dashed")+
  geom_line(aes(y=Ef+2*sigma), color=set1[1],linetype="dashed")

```

plot posterior

```{r}
subset(draws1, variable="f") %>%
  thin_draws(thin=5)%>%
  as_draws_df() %>%
  pivot_longer(!starts_with("."),
               names_to="ind",
               names_transform = list(ind = readr::parse_number),
               values_to="mu") %>%
  mutate(time=omicron_date_perc$num_week[ind])%>%
  ggplot(aes(time, mu, group = .draw)) +
  geom_line(color=set1[2], alpha = 0.1) +
  geom_point(data=omicron_date_perc, mapping=aes(x=num_week,y=perc_omicron), inherit.aes=FALSE)+
  geom_line(data=cbind(omicron_date_perc,pred), mapping=aes(x=num_week,y=Ef), inherit.aes=FALSE, color=set1[1], size=1)+
  labs(x="Week (numeric)", y="Percent Omicron")
```

